---
title: "DATA ANALYSIS AND VISUALIZATION IN R"
author: "Clifford Oduor,MSc"
date: "`r Sys.Date()`"
output: 
  html_document: 
    theme: journal
    highlight: tango
    toc: true
---

# **Introduction to R**

R is a system for data manipulation, statistical computation and graphics liked for its flexibility in
data manipulation, high-level graphics, and numerous specialized add-on packages that will
handle almost all - if not all - of your statistical needs.
The latest version of R can be downloaded from the Comprehensive R Archive Network- CRAN - main website is http://cran.r-project.org.

### *Important Note:*
1. R is case-sensitive
2. The # character at the beginning of a line signifies a comment, and is therefore not
executed. This is very helpful in describing codes and their functions
3. Help files for R functions are obtained by preceding the function name with ? (e.g. ?mean)

These training sessions have been organized into 4 main sessions.Each session
includes a short talk and a practical session during which participants are allowed sufficient
time for the more exciting interaction with R (perhaps the fastest route to being comfortable
working with R). The initial sessions will be spent familiarizing ourselves with the R environment,
installing/loading packages, creating and using R functions. Thereafter we will use R for data
wrangling, descriptives and elementary statistics, before delving into the more engaging
graphical display of data and the use of R for statistical modelling and Machine learning.
As with any programming language, investments made on mastering the basics are likely to pay off in the
advanced sessions.

### lets start coding 
```{r}
2+5
```

```{r}
#Operators < − or = are used to assign things or store objects
obj1<-2+5
obj1
```

```{r}
sessionInfo()
```

