---
title: "DATA SCIENCE IN R"
author: "Clifford Oduor,MSc"
date: "`r Sys.Date()`"
output: 
  html_document: 
    theme: journal
    highlight: tango
    toc: true
---

# **Introduction to R**

R is a system for data manipulation, statistical computation and graphics liked for its flexibility in
data manipulation, high-level graphics, and numerous specialized add-on packages that will
handle almost all - if not all - of your statistical needs.
The latest version of R can be downloaded from the Comprehensive R Archive Network- CRAN - main website is http://cran.r-project.org.

### *Important Note:*
1. R is case-sensitive
2. The # character at the beginning of a line signifies a comment, and is therefore not
executed. This is very helpful in describing codes and their functions
3. Help files for R functions are obtained by preceding the function name with ? (e.g. ?mean)
4. R resources and finding help
* stack overflow
* Rbloggers
* Rseek


These training sessions have been organized into 4 main sessions.Each session
includes a short talk and a practical session during which participants are allowed sufficient
time for the more exciting interaction with R (perhaps the fastest route to being comfortable
working with R). The initial sessions will be spent familiarizing ourselves with the R environment,
installing/loading packages, creating and using R functions. Thereafter we will use R for data
wrangling, descriptives and elementary statistics, before delving into the more engaging
graphical display of data and the use of R for statistical modelling and Machine learning.
As with any programming language, investments made on mastering the basics are likely to pay off in the
advanced sessions.

Now let's start coding 

### Data types
A dataset may contain variables with different data types.

```{r}
rm(list=ls())
gc()
```

```{r}
#Numeric
2+5
```

```{r}
#Character
#Operators < − or = are used to assign things or store objects
dat1 <- c("two","three","six","fifteen")
dat1

```

```{r}
#Operators < − or = are used to assign things or store objects
num1<-2+5
num1
```

```{r}
#Logical 
obj1<- TRUE
obj2<- FALSE
obj3<- obj1 & !obj2
obj3

```

```{r}
#Data structures
#vectors:these are one-dimensional array that can hold data of the same type 
dat2 <- c(2,3,6,15)
dat2
```


```{r}
#data frame:two-dimensional collection of elements of different data types 
dat3 <- data.frame(dat1,dat2)
dat3
```

```{r}
#checking if an object is  of type integer
is.integer(obj1) 
```

```{r}
#Check data type
typeof(obj1)

```
### Functions in R
R functions are a special type of object with
power to change or manipulate other objects.

```{r}
mean(dat3)

```
### Creating your own R function
+ They structure code, making it easier to read and maintain the code
+ They help in reusing of code, hence reducing duplication of code- Duplication of code is a bad programming style.
```{r}
divid <- function(x,y){
  result <- x/y;
  return(result)}

divid(12,8)
```
### *Exercise 1*
1. Create an object named m with the value 15.
2. Create an object named n with the value 17.
3. Add m to n.
4. Divide m by n.
5. Create a function called mulp() which multiplies two numbers.
6. Send 40 and 80 to mulp().


# **Importing data and Data wrangling in R**

Data read into R can be in two main formats. 
1. default R format *.rdata, and 
2. Text files mainly tab-delimited text files *.TXT, comma separated versions *.CSV 

Data may be imported into R using two ways:
1. Accessing datasets “built in” R which come as part of a libraries loaded into the console
2. Importing data from external sources (eg Excel spreadsheets, Access databases,MySQL, text files, spss files, sas files…etc)

### *R Packages:*
R Packages are collections of R functions, data, and compiled code in a well-defined format.
* The directory where packages are stored is called the library. 
* R comes with a standard set of packages. 
* Others are available for download and installation. 
* Once installed, they have to be loaded into the session to be used.

### *Import*

We are going to import data on Maternal mortality ratios(MMR) and The Human Development Index (HDI) for all countries in the world.
HDI measures each country's social and economic development. 

### Installing and using R packages
```{r}
# To Add a Package use: 
#install.packages("dplyr")
#install.packages("haven")
#install.packages("Rtools")
#install.packages("table1")

require(foreign)
require(haven)
require(dplyr)
require(readxl)
require(readr)
require(rio)
require(ggplot2)
require(magrittr)
library(naniar)

```

```{r}
#To see all packages installed use: 
library()
```

```{r}
#To see packages currently loaded 
search()
```

```{r}
data1<-read.csv("C:/MSH/DATA_SCIENCE/MY_TRAINING/datasets/MMR/dataset1.csv",header = TRUE,sep=",",na="")
data2<-read.csv("C:/MSH/DATA_SCIENCE/MY_TRAINING/datasets/MMR/dataset2.csv",header = TRUE,sep=",",na="")
```

### Viewing data
```{r}
head(data1,5)
```

```{r}
tail(data1,5)
```


```{r}
dim(data1)
```

```{r}
View(data1)
```

```{r}
#view data type
str(data1)
```

```{r}
names(data1)

```

Once you have imported your data set in r, in order to use it for analysis you can either attach it to r using attach() command or use $ sign to point to a variable a data frame.

```{r}
summary(data1$HDI_Rank)
```

```{r}
names(data1)
```

```{r}
names(data2)
```

###Checking for inconsistencies
```{r}
miss_var_summary(data1,order=TRUE)
```

```{r}
miss_var_summary(data2,order=TRUE)
```

###Handling missing values 
```{r}
View(data1[is.na(data1$HDI_Rank),])

```

```{r}
View(data2[is.na(data2$MMR),])

```


```{r}
# After reviewing the data we decide to replace the missing values for id 4:
data2$MMR[data2$ID==4]=45

```


```{r}
#replacing missing values with mean
data2$MMR[is.na(data2$MMR)] <- round(mean(data2$MMR, na.rm = TRUE))
data1$HDI_Rank[is.na(data1$HDI_Rank)] <- round(mean(data1$HDI_Rank, na.rm = TRUE))

```


```{r}
#Identify correct value from the questionnaires
table(data1$Continent)
```

```{r}
#Make appropriate changes using the command:
data1$Continent[data1$Continent=="AAfrica"]<-"Africa"

```

```{r}
#checking for duplicates
dup<-duplicated(data1$ID)
table(dup)
```


```{r}
#Transform continuous variable to categorical variables
data1$hdicat<-NA
data1$hdicat[data1$HDI_Rank<=49]=1
data1$hdicat[data1$HDI_Rank>49]=2
#labeling
data1$hdicat=factor(data1$hdicat,levels=c(1,2),labels=c("<Below 50","50 and Above"))
table(data1$hdicat)
```

```{r}
#convert character to numeric
table(data1$Hemisphere)
```

```{r}
data1$Hemispherecat<-ifelse(data1$Hemisphere=="Northern Hemisphere",1,2)
table(data1$Hemispherecat)
```


```{r}
data1$Hemispherecat=factor(data1$Hemispherecat,levels=c(1,2),labels=c("North","south"))
table(data1$Hemispherecat)
```

###Adding variables

```{r}
#rnorm(891, mean = 37, sd = 2) 

#temp<-rnorm(186,37,2)
```

```{r}
#datab<-cbind(data1,temp)
#head(datab)

```
###Appending and merging data frames 
When appending two datasets, the names of the variables must be the same
```{r}
#install.packages("gtools")
require(gtools)
```

```{r}
data3<-read.csv("C:/MSH/DATA_SCIENCE/MY_TRAINING/datasets/MMR/dataset3.csv",header = TRUE,sep=",",na="")
```


```{r}
#Transform continuous variable to categorical variables
data3$hdicat<-NA
data3$hdicat[data3$HDI_Rank<=49]=1
data3$hdicat[data3$HDI_Rank>49]=2
#labeling
data3$hdicat=factor(data3$hdicat,levels=c(1,2),labels=c("<Below 50","50 and Above"))
table(data3$hdicat)

```

```{r}
data3$Hemispherecat<-ifelse(data3$Hemisphere=="Northern Hemisphere",1,2)
table(data3$Hemispherecat)
```

```{r}
data3$Hemispherecat=factor(data3$Hemispherecat,levels=c(1,2),labels=c("North","south"))
table(data3$Hemispherecat)
```

```{r}
#appending two datasets
data_append<-smartbind(data1,data3)
```

```{r}
#merging two datasets 
merged_data<-merge(data_append,data2,by=c("ID"))
View(merged_data)
#merged_data<- merge(data1,data2,by.x="PassengerId",by.y="PassengerId", all.x=T)
```
###subsetting data frames 
```{r}
data_Asia<-merged_data %>%
filter(Continent=="Asia")
```

```{r}
#Subsetting data
data_Asia_100<- subset(merged_data, Continent =="Asia" & MMR<100)
```


###stack and unstack data frames (long/wide format)
```{r}
#stack_df<-stack(data_males)

```

```{r}
#unstack_df<-unstack(stack_df)

```


```{r}
#Saving data
save(merged_data, file = "C:/MSH/DATA_SCIENCE/MY_TRAINING/datasets/MMR/merged_data.Rdata")

```

#**Exploratory data analysis (EDA)**

```{r}
rm(list=ls())
load(file = "C:/MSH/DATA_SCIENCE/MY_TRAINING/datasets/MMR/merged_data.Rdata")

```

```{r}
head(merged_data,2)
```

```{r}
dim(merged_data)
```


```{r}
names(merged_data)

```

```{r}
#frequency tables
table(merged_data$Hemispherecat)
```

```{r}
#Table of proportions
prop.table(table(merged_data$Hemispherecat))*100

```

```{r}
#install.packages("Rtools")
#install.packages("table1")
#require (table1)

```

```{r}
#table1(~factor(data1$sexcat))

```

```{r}
#Cross tabulations of categorical variables

table(merged_data$Hemispherecat,merged_data$Continent)
```


```{r}
#install.packages("shiny")
#require(shiny)
```

```{r}
#set up factors
# merged_data$Survived<-as.factor(merged_data$Survived)
# merged_data$Sex<-as.factor(merged_data$Sex)
# merged_data$agecat1<-as.factor(merged_data$agecat1)
# merged_data$Embarked<-as.factor(merged_data$Embarked)
```

```{r}
#able1(~(Sex)+(agecat1)+(Embarked)|Survived, data = merged_data)

```

```{r}
#Descriptive statistics for continuous variables
summary(merged_data$MMR)
```
###Data Visualization 

```{r}
library(ggplot2)
```
 
```{r}
str(merged_data)
```
 
 
```{r}
merged_data$MMR<-as.numeric(merged_data$MMR)
```
 

```{r}
#bar graph
ggplot(data=merged_data, aes(x=Hemispherecat,y=MMR,fill=Hemispherecat))+ 
         geom_bar(stat="identity")+ylim(0,1000)+
         ggtitle("Bar graph of MMR by Hemisphere")

```

```{r}
# #Add new variable say dob
# dob<-sample(seq(as.Date('1944/01/01'), as.Date('2024/01/01'), by="day"), 891)
# data4<-cbind(data3,dob)

```

```{r}
#Box plot
box1<- ggplot(data=merged_data,aes(x=Hemispherecat, y=MMR,fill=Continent)) + geom_boxplot() + ggtitle("Distribution of MMR across continent")+ ylab("MMR")
box1

```

```{r}
#Histogram
hist1<-ggplot(data=merged_data,aes(x=MMR))+  geom_histogram()+ggtitle("A histogram of MMR")+xlab("MMR")+ylab("Frequency")
hist1

```

```{r}
#scatter plot
plot<-ggplot(merged_data, aes(HDI_Rank, MMR)) + geom_point() + xlim(0, 200)+ ylim(0,1000)
plot

```


```{r}
#How to convert ggplot2 graph into a plotly
#install.packages("plotly")
library(plotly)
ggplotly(plot)

```


```{r}
ggplot(data=merged_data,aes(x=HDI_Rank,y=MMR)) + geom_point()+geom_smooth(method = "lm")+
  scale_x_continuous("HDI")+ scale_y_continuous("MMR")
```

```{r}
#Line plot with multiple groups

# lp<-ggplot(data=merged_data, aes(x=Age, y=Fare, group=Embarked)) +
#   geom_line()
# lp

```

```{r}
names(merged_data)
```

# **Elementary statistics**

```{r}
## Visualise the distribution of age
##Checking for normality
hist(merged_data$MMR, xlab="MMR", col="green",main="")
```

```{r}
hist(merged_data$HDI_Rank, xlab="HDI_Rank", col="green",main="")
```

```{r}
#Shapiro-Wilk test can be performed as follow:
#Null hypothesis: the data are normally distributed
shapiro.test(merged_data$MMR) 
shapiro.test(merged_data$HDI_Rank) 
```

```{r}
#shapiro.test(merged_data$Fare)
```


```{r}
## Testing for normality
# qqnorm(data1$Age)
# qqline(data1$Age,lty=2)

```

```{r}
names(merged_data)
```

```{r}
plot(merged_data$HDI_Rank,merged_data$MMR) 
```

### Spearman rank correlation coefficient
```{r}
#measures a linear dependence between two variables (x and y)
cor(merged_data$HDI_Rank, merged_data$MMR,  method = "spearman", use = "complete.obs")

```

### Two-sample t-test for means (comparing two independent means)
```{r}
#Null hypothesis: there is no difference in MMR by North and south hemisphere
t.test(merged_data$MMR~as.factor(merged_data$Hemispherecat), data=merged_data)

```

```{r}
#One-way ANOVA
#Used to determine variation between different groups of the independent variable
anova(lm(merged_data$MMR~as.factor(merged_data$Continent)))
```

### Chi-square test
```{r}
#cross tabulation
#Suppose in our data we wanted to establish if test differed by sex
table(merged_data$sexcat,merged_data$Test.Results)
```

```{r}
## conduct a chisq test
chisq.test(merged_data$hdicat,merged_data$Hemispherecat)
```

# **Statistical modelling in R**
### Purpose of modelling is understanding relationships between variables and allows us to make predictions on previously unseen data
```{r}
library(ISLR2)
library(MASS)
library(caret)
library(splines)
library(pROC)
```

```{r}
rm(list=ls())
load(file = "C:/MSH/DATA_SCIENCE/MY_TRAINING/datasets/MMR/merged_data.Rdata")
```

### Linear regression

```{r}
plot(merged_data$MMR, merged_data$HDI_Rank)
```

```{r}
#What is the association between age of patients and billing amount?
model1<-lm(MMR~HDI_Rank,data=merged_data)
summary(model1)

```
### Evaluating model accuracy
### Root mean square error: measures the average difference between values predicted by a model and the actual values.
### It provides an estimation of how well the model is able to predict the target value (accuracy). 
### The lower the value of the Root Mean Squared Error, the better the model is.

```{r}
mse =mean((merged_data$MMR-predict(model1))^2)
rmse<- sqrt(mean((predict(model1)-merged_data$MMR)^2))
mse 
rmse
```


```{r}
###Saving the Model#####
saveRDS(model1, file="C:/MSH/DATA_SCIENCE/MY_TRAINING/datasets/MMR/model1.RDS")
#model<-readRDS("C:/MSH/DATA_SCIENCE/MY_TRAINING/datasets/MMR/model1.RDS")

```

### Multivariable linear regression
```{r}
#What is the association between age,gender of patients and billing amount?
model2<-lm(MMR~HDI_Rank+Hemispherecat,data=merged_data)
summary(model2)

```

```{r}
str(merged_data$hdicat)
```

### Logistic(Binary) regression
```{r}
model3<-glm(hdicat~MMR,data=merged_data,family="binomial")
summary(model3)
summary(model3)$coef

```

```{r}
exp(coef(model3))
```

```{r}
exp(cbind(Odds_Ratio = coef(model3), confint(model3)))
```

# **MACHINE LEARNING in R**

#From Previous lessons we have:
Explored the data
Data cleaning
Prepared

Pick learning:predict a countrie HDI category based on different predictors/features/variables/columns
```{r}
library(naniar)     # used for checking for missing values
library(dplyr)      # used for data manipulation and joining
library(ggplot2)    # used for plotting 
library(caret)      # used for modeling
library(cowplot)    # used for combining multiple plots 

```


```{r}
rm(list=ls())
load(file = "C:/MSH/DATA_SCIENCE/MY_TRAINING/datasets/MMR/merged_data.Rdata")
```
### *Explore data*
```{r}
#EXPLORE DATA
miss_var_summary(merged_data,order=TRUE)
```

```{r}
gg_miss_var(merged_data)
```

```{r}
names(merged_data)
```

### *Feature Engineering*
#Feature engineering entails reformatting predictor values to make them easier for a model to use effectively
Involves:Converting variables into factors,Selecting specific columns,filter rows based on certain criteria,group data by one or more variables and perform aggregate calculations,
reshape and restructure data,normalization, scaling, encoding, and feature selection.
It is a crucial step in building a good machine-learning model.Helps increase the accuracy of the model thereby enhancing the results of the predictions.

```{r}
#Selecting specific columns
subset_data<-subset(merged_data,select =-c(ID,HDI_Rank,Hemisphere))
```

```{r}
str(subset_data)
```

```{r}
#one hot encoding of specific variables
oneHot_vars<-subset(subset_data,select=c(Continent))
dummy <- dummyVars(" ~ .", data=oneHot_vars)
newdata <- data.frame(predict(dummy, newdata = oneHot_vars))
```

```{r}
hot<-subset(subset_data,select=-c(Continent))
```

```{r}
subset_data1<-cbind(hot,newdata)
miss_var_summary(subset_data1,order=TRUE)
```

```{r}
#SPLITTING THE DATA FOR TRAINING AND LOCAL TESTING
indexes<-createDataPartition(subset_data1$hdicat,times=1,p=0.8,list =FALSE)

train<-subset_data1[indexes,]
test<-subset_data1[-indexes,]
```

### *Algorithm selection*























```{r}
sessionInfo()
```

